---
title: "statistical analysis"
output:
  html_document:
    df_print: paged
---

Objective: Exploring the trends and factors related to suicide rates across different countries, years, genders, and age groups.

#Loading data

```{r}
library(tidyverse)
library(gganimate)
library(ggplot2)
library(tidyr)
data<-read.csv("who_suicide_statistics.csv")
str(data)
summary(data)
```

## Handling null values

```{r}
# imputing data
data_imputed <- replace_na(data, list(suicides_no = 0,population=0))
```

# Descriptive statistics

```{r}
# Summary statistics of suicide numbers and population grouped by country
summary_stats <- data_imputed %>%
  group_by(country) %>%
  summarize(
    mean_suicides = mean(suicides_no),
    mean_population = mean(population)
  )

# Displaying summary statistics
head(summary_stats)
# Summary statistics of suicide numbers and population grouped by year
summary_stats <- data_imputed %>%
  group_by(year) %>%
  summarize(
    mean_suicides = mean(suicides_no),
    mean_population = mean(population)
  )

# Displaying summary statistics
head(summary_stats)
# Summary statistics of suicide numbers and population grouped by age
summary_stats <- data_imputed %>%
  group_by(age) %>%
  summarize(
    mean_suicides = mean(suicides_no),
    mean_population = mean(population)
  )

# Displaying summary statistics
head(summary_stats)
# Summary statistics of suicide numbers and population grouped by sex
summary_stats <- data_imputed %>%
  group_by(sex) %>%
  summarize(
    mean_suicides = mean(suicides_no),
    mean_population = mean(population)
  )

# Displaying summary statistics
head(summary_stats)
```

# Exploratory Data analysis

```{r}
data_imputed%>%
  group_by(country)%>%
  summarise(sum=sum(suicides_no))%>%
  slice_max(sum,n=10)%>%
  ggplot(aes(x=reorder(country,desc(sum)),y=sum))+geom_col()+labs(title="Total Suicide numbers according to country(top 10)")+theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
new<-data_imputed%>%
  # selecting required columns
  select(year,age,suicides_no)%>%
  group_by(year,age)%>%
  summarize(sum=sum(suicides_no))
# creating a ggplot
a <- ggplot(new, aes(x =age,y = sum,fill=age)) +
  geom_bar(stat = "identity") +
  labs(title = "Year: {closest_state}") +
  geom_text(aes(label = sum),
            position = position_dodge(0.9), vjust = -1 ) +
  ylim(0,100500)+
  theme_light()+
  transition_states(states = year, transition_length = 10000000000000, state_length = 1000000000000000) + 
  enter_fade() + 
  exit_shrink() +
  ease_aes('quadratic-in-out')+labs(title="time-series graph for trends in suicide number for each age group")
# gif 
a_gif <- animate(a, width = 940, height = 480)

a_gif
```

```{r}
data_imputed %>%
  # filtering data
  filter(population != 0) %>%
  select(country, suicides_no, population,year) %>%
  group_by(country) %>%
  summarise(s = sum(suicides_no), p = sum(population))%>%
  mutate(rate=s/p*100000)%>%
  slice_max(rate,n=10)%>%
  # creating plot
  ggplot(aes(x=country,y=rate,fill=country))+geom_col()+labs(title="suicide rate for top 10 countries")+theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Boxplot of suicides_no by sex
ggplot(data_imputed, aes(x = sex, y = suicides_no, fill = sex)) +
  geom_boxplot() +
  labs(title = "Number of Suicides by Sex", x = "Sex", y = "Number of Suicides") +
  theme_minimal()

```

# Regression analysis

```{r}
lm_model <- lm(suicides_no ~ year + sex + age, data = data_imputed)

# Summarizing the regression model
summary(lm_model)
```
> The coefficients indicate the estimated effect of each predictor variable on the number of suicides. For example, being male (sexmale) is associated with an increase of 192.89 in the number of suicides compared to being female. A p-value less than 0.05 (commonly used threshold) suggests that the variable has a significant effect on the number of suicides. The adjusted R-squared value (0.04025) indicates that the model explains about 4.025% of the variance in the number of suicides, suggesting that the model might not be a good fit for the data.

>The regression model suggests that variables such as sex and age group have a significant impact on the number of suicides, while the year variable and the intercept may not be statistically significant predictors.

# ANOVA analysis
```{r}
# Fitting ANOVA model
anova_model <- aov(suicides_no ~ sex, data = data_imputed)

# Summarizing ANOVA results
summary(anova_model)

```
>The ANOVA results indicate that there is a significant difference in the mean number of suicidesbetween males and females. The F-value is 678.1, and the p-value is less than 2e-16, which is highly significant.
